Appendix: A fragile guardrail---The minimal effect of election fraud on protest intensity in non-democracies

### 2025-11-17 Output/authoritarian_only_pdf_pop_rev.Rmd
Final edits to conclusion

There are competing views on the relationship between election fraud and protest. In one family of models, which I have called the regime-weakness model, incumbent's may hold cleaner elections when the risk of post-election protest is higher. Other models contend that election fraud can indicate incumbent strength. Building on the principal-agent literature of election fraud, I argue that large-scale fraud draws on the same resources that make incumbents resilient to protest, even when regime popularity (and thus vote-shares) are low. Conducting election fraud is made easier when incumbents have control over patronage resources and the ability to protect front-line agents from the consequences of illegal behavior; this capacity also helps insulate incumbents from the effects of mass protest. Resolving the tension between these views in the literature has important implications for the study of election fraud and authoritarian stability.

This paper attempts to test the assumptions of these two models by investigating the relationship between election fraud and two protest variables: initiation and duration. Specifically, both the regime-weakness and regime-strength models may make the same predictions regarding protest onset; their predictions differ with regard to protest duration. Drawing on data from the V-Dem, NELDA, and ECAV datasets, it uses cross-national data on 659 elections in electoral authoritarian regimes to show that protest occurrence is most likely when incumbent vote-shares are very low and election fraud is high, as both models predict. However, and most importantly, incumbents that engage in larger-scale election fraud are able to more quickly defeat protest movements.

@@

# TEst stuff

```{r include=FALSE, eval=FALSE}
library(lavaan)
library(dplyr)

# Prepare country-year level data
nelda2.sub <- nelda2.sub %>%
  mutate(
    # For country-years with protests: inverse of mean duration
    # Higher values = shorter duration = stronger regime
    protest_duration_inverse = if_else(
      any.ecav.protest == 1,
      1 / d.events.num,  # or -log(mean_duration)
      NA_real_
    )
  )

nelda2.sub <- nelda2.sub %>%
  mutate(taxstate.num = as.numeric(taxstate))
nelda2.sub <- nelda2.sub %>%
  mutate(rentierstate.num = as.numeric(rentierstate))
nelda2.sub <- nelda2.sub %>%
  mutate(unscheduled_election.num = as.numeric(unscheduled_election))

# MIMIC model
mimic_spec <- '
  # Latent variable
  regime_strength =~  gdpgro.1lag +v2elirreg.inv +   physinteg.1lag + urban.pct + rentierstate.num + taxstate.num +  unscheduled_election.num + inc.termlimit.num + winner.share.upd

  # Structural part: causes of regime strength
  protest_duration_inverse ~ regime_strength  + winner.share.upd

  # Allow correlation between indicators
 # protest_duration_inverse ~~ any.ecav.protest
 v2elirreg.inv ~~ physinteg.1lag 
 v2elirreg.inv ~~ urban.pct
 v2elirreg.inv ~~  inc.termlimit.num
'

# Fit with FIML to handle missing protest_duration_inverse
fit <- sem(mimic_spec, 
           data = nelda2.sub %>% filter(d.events.num >0 & inc.lose == 0),
           missing = "fiml")  # Full Information Maximum Likelihood

summary(fit, fit.measures = TRUE, standardized = TRUE)

# Extract latent scores
regime_scores <- lavPredict(fit, type = "lv")
data_cy$regime_strength_estimated <- regime_scores[, 1]
```

```{r eval=FALSE}
# Step 1: Create a composite measure of regime strength from protests
# (for country-years with protests)

nelda2.sub <- nelda2.sub %>%
  mutate(
    # Composite indicator (you can refine this)
    strength_indicator = case_when(
      any.ecav.protest == 0 ~ NA_real_,  # Missing for no protests
      any.ecav.protest == 1 ~ -log(d.events.num) # Lower duration = higher strength
    )
  )

# Step 2: SEM with measurement error correction
sem_spec <- '
  # Latent variable with single indicator
  # (need to fix measurement error variance)
  regime_strength =~ 1*strength_indicator

  # Estimate measurement error (you can set this based on reliability)
  strength_indicator ~~ error_var*strength_indicator

  # Structural model
  regime_strength ~  v2elirreg.inv + gdpgro.1lag +   physinteg.1lag
'

# Fit model
fit2 <- sem(sem_spec, 
            data = nelda2.sub,
            missing = "fiml")

summary(fit2, standardized = TRUE)
```
```{r eval=FALSE}
# This is cleaner and avoids identification issues
path_model <- '
  # Protest occurrence predicted by covariates
  any.ecav.protest ~ v2elirreg.inv + gdpgro.1lag +   physinteg.1lag

  # Protest duration (inverse) predicted by same covariates
  # Only estimated for observations with protests
  protest_duration_inverse ~ v2elirreg.inv + gdpgro.1lag +   physinteg.1lag

  # Allow correlation between equations
  any.ecav.protest ~~ protest_duration_inverse
'

fit_path <- sem(path_model, 
                data = nelda2.sub,
                missing = "fiml")

summary(fit_path, standardized = TRUE, fit.measures = TRUE)
```

```{r eval=FALSE}
# Use only country-years WITH protests
data_protests_only <- nelda2.sub %>%
  filter(any.ecav.protest == 1)

# Simple regression path model
protest_model <- '
  # Duration as function of covariates
  protest_duration_inverse ~ legcon.1lag + altinf.1lag +  osorg.1lag + gdpgro.1lag +  winner.share.upd  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate +  unscheduled_election + regional_success_lag + v2elirreg.inv

  # If you have multiple protests per country-year, 
  # you might want to include number of protests as control
  protest_duration_inverse ~ n.events.post
'

fit_protests <- sem(protest_model,
                    data = data_protests_only)

summary(fit_protests, standardized = TRUE, rsquare = TRUE)

# This tells you:
# 1. Which covariates predict shorter duration (stronger regimes)
# 2. How much variance in duration is explained
# 3. Specific effect of election fraud on regime strength
```

Structural equation modelling doesn't seem the way to go

```{r eval=FALSE}

df <- nelda2.sub %>% filter(inc.replaced.prot == 0) %>% dplyr::select(legcon.1lag, altinf.1lag,  osorg.1lag, gdpgro.1lag, inc.termlimit.num,  physinteg.1lag, physintlag.sq, urban.pct, rentierstate.num, taxstate.num ,  unscheduled_election.num, inc.termlimit.num, v2elirreg.inv, NeldaID)

df <- df %>% filter(complete.cases(.))

CFA.model <- ' regime_strength  =~  v2elirreg.inv + physinteg.1lag + physintlag.sq + altinf.1lag + osorg.1lag + legcon.1lag   + gdpgro.1lag + inc.termlimit.num  +  urban.pct + rentierstate.num + taxstate.num +  unscheduled_election.num + inc.termlimit.num'

fit <- cfa(CFA.model, data = df, std.lv=TRUE)
summary(fit)

regime_scores <- lavPredict(fit)

df <- df %>% mutate(rgm_st = regime_scores[,1])

df <- df %>% left_join(nelda2.sub, by = "NeldaID")

init_probit <- glm(any.ecav.protest ~ rgm_st*winner.share.upd + presidential, binomial(link = "probit"), data = df)
df_probit <- init_probit$data
df_probit <- df_probit %>%  dplyr::select(legcon.1lag.x, altinf.1lag.x,  osorg.1lag.x, gdpgro.1lag.x, inc.termlimit.num.x,  physinteg.1lag.x, physintlag.sq.x, urban.pct.x, rentierstate.num.x, taxstate.num.x ,  unscheduled_election.num.x, inc.termlimit.num.x, v2elirreg.inv.x, NeldaID, winner.share.upd, d.events.num, rgm_st)
df_probit <- df_probit %>% filter(complete.cases(.))

df_probit$IMR <- sampleSelection::invMillsRatio( init_probit )$IMR1

library(survival)
library(ggsurvfit)

df <- df %>% mutate(surv_status = ifelse(d.events.num > 0, 1, NA))
df_probit <- df_probit %>% mutate(surv_status = ifelse(inc.replaced.prot == 1, 0, surv_status))
df_probit <- df_probit %>% mutate(surv_status_censor_assume = ifelse(d.events.num == 0, NA, 
                                                                       ifelse(d.events.num >= 69, 0, 1)))

cox1 <- coxph(Surv(d.events.num, surv_status) ~  rgm_st*winner.share.upd  +  median.participants + regional_success_lag, data = df)
summary(cox1)

## Unclear what's going on here. But the general idea is that election fraud negatively loads onto the 'regime strength' variable, which in turn has no apparent effect on duration.
```

```{=latex}
\begin{table}[]
\begin{tabular}{|l|l|l|l|l|}
\hline
\multicolumn{1}{|r|}{Protest is assumed to be:} &
  \multicolumn{1}{c|}{Instrumental protest} &
   &
  \multicolumn{1}{c|}{Multipurpose protest} &
   \\ \hline
\multicolumn{1}{|r|}{Incumbent's genuine popularity:} & Low & High & Low & High \\ \hline
Manipulation...                                       &     &      &     &      \\ \hline
...obscures strength or indicates weakness &
  (I) High and increasing &
  (I): Low and increasing &
  (I): High and increasing &
  (I): Low and increasing \\ \hline
...reflects strength &
  (I): High and decreasing &
  (I): Low and decreasing &
  (I): High and increasing &
  (I): Low and no effect \\ \hline
\end{tabular}
\caption{}
\label{tab:my-table}
\end{table}
```

>>>>

There are competing views on the relationship between election fraud and protest. In one family of models, which I have called the regime-weakness model, incumbents may hold cleaner elections when the risk of post-election protest is higher. Other models contend that election fraud can indicate incumbent strength. Building on the principal-agent literature of election fraud, I argue that large-scale fraud draws on the same resources that make incumbents resilient to protest, even when regime popularity (and thus vote-shares) are low. Conducting election fraud is made easier when incumbents have control over patronage resources and the ability to protect front-line agents from the consequences of illegal behavior; this capacity also helps insulate incumbents from the effects of mass protest. Resolving the tension between these views in the literature has important implications for the study of election fraud and authoritarian stability.

This paper tests the assumptions of these two models by investigating the relationship between election fraud and the duration of protest. Though both models may make the same predictions regarding protest onset; their predictions differ with regard to protest duration. Drawing on data from the V-Dem, NELDA, and ECAV datasets, it uses cross-national data on 659 elections in electoral authoritarian regimes to show that incumbents that engage in larger-scale election fraud are able to more quickly defeat protest movements.

@@
